---
title: "create MAE"
output: html_document
---

```{r}
if (!requireNamespace('BiocManager'))
  install.packages('BiocManager')

if (!requireNamespace('MultiAssayExperiment'))
  BiocManager::install('MultiAssayExperiment')

if (!requireNamespace('data.table'))
  install.packages('data.table')
```
```{r}
library(MultiAssayExperiment)
library(data.table)
library(magrittr)
```

Helper functions:

```{r}
## functions to get number of common samples/features for two datasets
scom <- function(x,y) sum(colnames(x) %in% colnames(y))
fcom <- function(x,y) sum(rownames(x) %in% rownames(y))

## functions to get names of mismatching samples for two datasets
smismatch <- function(x,y) colnames(x)[!colnames(x) %in% colnames(y)]

## pairwise common samples for a list of datasets
scom_list <- function(data_list1, data_list2 = data_list1) sapply(data_list1, function(x) sapply(data_list2, function(y) scom(x, y)))

## pairwise common feartures for a list of datasets
fcom_list <- function(data_list1, data_list2 = data_list1) sapply(data_list1, function(x) sapply(data_list2, function(y)fcom(x, y)))

name_mismatch <- function(data_list1, data_list2=data_list1) lapply(data_list1, function(x) lapply(data_list2, function(y)smismatch(x, y)))

## function to read tsv files into data.frames with first column as rownames
fread_tsv <- function(file_list) {
  out <- lapply(file_list, function(x) as.matrix(data.frame(fread(x), row.names = 1)))
  ## name the list with file name
  names(out) <- gsub(".tsv", "", basename(file_list))
  out
}
```

# To DO

----
* Why are there no metadata for metabolomics data?
* See why metadata is missing a lot of sample names in WGS data, and why there are duplicates

----

# Data

File names:

```{r}
## input
metab_files <- list.files(path = '../data/merged_Metabolomics/', full.names = TRUE)
wgs_files <- list.files(path = '../data/Merged_WGS/', full.names = TRUE)
metadata_file <- '../data/metadata.tsv'
prot_files <- list.files('../data/Proteomics/', full.names = TRUE)
```

Read files:

## metadata

```{r}
metadata <- data.frame(fread(metadata_file))
```

```{r}
##TODO Why are there duplicate metadata
## duplicate metadata
sum(duplicated(metadata))
```


## Metabolomic

```{r}
## read files as named list
metab_list <- fread_tsv(metab_files)
```

```{r}
## unintended X added to colnames with subject_id that start with numbers
lapply(metab_list, function(x) grep(pattern = "^X", colnames(x)))
## drop it
metab_list <- lapply(metab_list, function(x) set_colnames(x, gsub(pattern = "^X",replacement = "", colnames(x))))
```

Check all samples have metadata:

```{r}
##TODO There are no metadata available for these samples
## number of samples in each dataset not present in metadata$new_ID
lapply(metab_list, function(x) sum(! colnames(x) %in% metadata$new_ID))

## number of trimmed sample names (no visit) in each dataset not present in metadata$sample_id
lapply(metab_list, function(x) sum(! gsub(pattern = "_.*$", "", x = colnames(x)) %in% metadata$sample_id))
## so all samples are valid, only their visit record is not there
```

```{r}
View(metab_list[[1]][1:10,1:10])
```

<!-- [#Antoine: this dataset is using the sample_id, do we need the new_ID? We can use different names for samples across assays and create a sampleMap that defines it, but is this necessary or should we just harmonise them all?] -->

```{r}
## check that new ID is same as subjectid_visitnumber
sum(metadata$new_ID != paste0(metadata$subject_id, "_", metadata$visit_number))
```

```{r}
## common samples
scom_list(metab_list)
## common features
fcom_list(metab_list)
```

```{r}
## adding 'metab_' to names to distinguish in MAE object
names(metab_list) %<>% paste0("metab_", .)
```


## WGS

Same with WGS data:

```{r}
## read files as named list
wgs_list <- fread_tsv(wgs_files)
```

```{r}
## unintended X added to colnames with subject_id that start with numbers
lapply(wgs_list, function(x) grep(pattern = "^X", colnames(x)))
## drop it
wgs_list <- lapply(wgs_list, function(x) set_colnames(x, gsub(pattern = "^X",replacement = "", colnames(x))))
```

```{r}
## [#Antoine: why are the column names (which I assume are supposed to be new_ID) sometimes using sample_id and sometimes subject_id?]
View(wgs_list[[3]][1:10,1:10])
```

```{r}
## common samples
scom_list(wgs_list)
## common features
fcom_list(wgs_list)
```

What are the non-matching ones?

```{r}
wgs_mismatch <- name_mismatch(wgs_list)
```

```{r}
## one-by-one mismatches can be inspected as follows:
## for pathcoverage and pathabundance
wgs_mismatch$pathcoverage$pathabundance
## this suggests there's been 2 NA's in headers of one dataset

wgs_mismatch$pathcoverage$WGS_community
## this suggests there's been 2 NA's in headers of one dataset, plus an extra sample
## which does not exist in metadata:
"1419f08f554e0c93f3b62fe90c246a08_13" %in% metadata$new_ID
```

The main mismatch comes from ecs (~210 samples):

```{r}
## those from genefamilies not present in ecs
head(wgs_mismatch$genefamilies$ecs)
```

 Are these valid new_ID's?
 
```{r}
sum(wgs_mismatch$genefamilies$ecs %in% metadata$new_ID)
```

Doesn't look like it


```{r}
## number of genefamilies samples present in metadata
sum(colnames(wgs_list$genefamilies) %in% metadata$new_ID)
```

[#Antoine: there seems to be a lot of samples in WGS dataset with no metadata]


Shared samples with metabolites:

```{r}
## common samples b/w WGS (~300) and metabolomics (545)
scom_list(wgs_list, metab_list)
```

[#Antoine] we noticed that there were `NA`s in original files that lead to weird namings by `fread`. Also you can see the `ecs` data does not share a lot of samples with others. Can you please let me know if I have to i) drop `ecs` dataset and ii) drop some non-matching samples from other datasets?

I'm assuming these are separate data so I don't worry about common features.

```{r}
## adding 'WGS_' to names to distinguish in MAE object
## exclude WGS_community
names(wgs_list) <- gsub(".tsv", "", basename(wgs_files))
names(wgs_list) %<>% paste0("WGS_", .)
## last one does already have WGS prefix
names(wgs_list)[5] <- gsub(".tsv", "", basename(wgs_files)[5])
names(wgs_list)
```


## Proteomic

```{r}
## read files as named list
prot_list <- fread_tsv(prot_files)
```

```{r}
## this one seems to use new_ID = paste0(sample_id,"_",visit)?
View(prot_list[[1]][1:10,1:10])
```

```{r}
## common samples
scom_list(prot_list)
## common features
fcom_list(prot_list)
```

[#Antoine] So I'm going with the one from `pep` ones. Also keeping the `kos` one as separate dataset, right?

```{r}
## keeping 2pep_5p_FDR_5ppm and kos datasets
prot_list <- prot_list[4:5]

## adding 'prot_' to names to distinguish in MAE object
names(prot_list) %<>% paste0("prot_", .)
```

Create MAE object:

```{r}
iHMP <- MultiAssayExperiment(experiments = ExperimentList(c(prot_list, wgs_list, metab_list))
                     #, colData = data.frame(metadata, row.names = "new_ID") ## this should use a row.name that matches the sample names - not possible at the moment because there are a lot of duplicated and we don't know which ones we wanna use
                     # as sample names
                     )
```


